---
title: "Medication Adherence:  Independent Investigation"
author: "Rohit Shivthare"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results='asis', fig.align='center', fig.cap = TRUE)
```

```{r seed}
set.seed(41)
```

```{r libraries, warning =FALSE, message= FALSE}
library(data.table)
library(DT)
library(lmtest)
library(dplyr)
library(ggplot2)
```

```{r constants}
# Adherence data columns
id.name <- "id"
t1.name <- "t1"
t2.name <- "t2"
ace.name <- "ace"
bb.name <- "bb"
statin.name <- "statin"

# Baseline Measurements columns
age.name <- "age"
gender.name <- "gender"
region.name <- "region"
diabetes.name <- "diabetes"
baseline.condition.name <- "baseline.condition"

# Days
one.year.days <- 365
two.weeks.days <- 14
```

```{r functions}
round.numerics <- function(x, digits){
  if(is.numeric(x)){
    x <- round(x = x, digits = digits)
  }
  return(x)
}
```

```{r load_data}
# Reading adherence and baseline measurements datasets
adherence = fread("../Data/adherence.csv")

baseline = fread("../Data/baseline measurements.csv")

# Merging baseline data with adherence
merged_data <- merge(adherence, baseline, by = id.name)
```

```{r explore_data, eval = FALSE}
# To understand the structure of both the datasets
str(adherence)
str(baseline)

# First 5 records for both the datasets
head(adherence)
head(baseline)

# Records for a random Patient ID
adherence[,id == "KQ68Ubs0yMY5ID5a",]

adherence %>%
  filter(id == "KQ68Ubs0yMY5ID5a") %>%
  select(id,t1,t2,ace,bb,statin)

# To know the unique values
unique(adherence$id)
unique(baseline$region)
unique(baseline$diabetes)
unique(baseline$gender)
unique(baseline$diabetes)
```
### Independent Analysis

```{r Filtering patient data for over a year}
# Getting the patients for over one year
patients_1_year <- adherence[,.SD[max(get(t2.name)) - min(get(t1.name)) > one.year.days], by=get(id.name)]

# Calculate medicine adherence
patients_1_year[, medicine_adherence_ace := sum(get(ace.name)) / (max(get(t2.name)) - min(get(t1.name))), by = get(id.name)]
patients_1_year[, medicine_adherence_bb := sum(get(bb.name)) / (max(get(t2.name)) - min(get(t1.name))), by = get(id.name)]
patients_1_year[, medicine_adherence_statin := sum(get(statin.name)) / (max(get(t2.name)) - min(get(t1.name))), by = get(id.name)]

# Merge adherence and baseline datasets on id.name
data_impact_adherence <- merge(baseline, patients_1_year[, .(id, medicine_adherence_ace, medicine_adherence_bb, medicine_adherence_statin)], by = id.name)

```

#### Independent Analysis Question 1 - Does age and gender have an impact on adherence rate of ACE Inhibitors, Beta Blockers and Statins?

```{r IP1}
# Fit linear regression model for adherence with age and gender
model_age_gender_ace <- lm(medicine_adherence_ace ~ age + gender, data = data_impact_adherence)
model_age_gender_bb <- lm(medicine_adherence_bb ~ age + gender, data = data_impact_adherence)
model_age_gender_statin <- lm(medicine_adherence_statin ~ age + gender, data = data_impact_adherence)

# Display model summary
summary_age_gender_ace <- summary(model_age_gender_ace)

summary_age_gender_bb <- summary(model_age_gender_bb)

summary_age_gender_statin <- summary(model_age_gender_statin)

# Create age groups
data_impact_adherence$age_group <- cut(data_impact_adherence$age, breaks = c(seq(30, 90, by = 20), Inf))
```

##### Bar plot for Ace Inhibitors adherence vs age group and gender

We can notice that Females tend to be more adherent in the yonger age group {30 - 50 years} while Males are more adherent in the elder age group 50+ onwards in each of the medications (ACE Inhibitors, Beta Blockers and Statins)

```{r bar_plot_age_gender_ace}
ggplot(data_impact_adherence, aes(x = age_group, y = medicine_adherence_ace, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Age Group", y = "Ace Adherence", title = "Ace Adherence vs Age Group and Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

##### Bar plot for Beta Blockers adherence vs age group and gender

```{r bar_plot_age_gender_bb}
ggplot(data_impact_adherence, aes(x = age_group, y = medicine_adherence_bb, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Age Group", y = "Bb Adherence", title = "Bb Adherence vs Age Group and Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels
```

##### Bar plot for Statin adherence vs age group and gender

```{r bar_plot_age_gender_statin}
ggplot(data_impact_adherence, aes(x = age_group, y = medicine_adherence_statin, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Age Group", y = "Statin Adherence", title = "Statin Adherence vs Age Group and Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels
```

#### Independent Analysis Question 2 - Does diabetes and baseline condition have an impact on adherence rate of ACE Inhibitors, Beta Blockers and Statins?

```{r IP2}
# Fit linear regression model for adherence with diabetes and baseline condition
model_diabetes_baseline_ace <- lm(medicine_adherence_ace ~ diabetes + baseline.condition, data = data_impact_adherence)
model_diabetes_baseline_bb <- lm(medicine_adherence_bb ~ diabetes + baseline.condition, data = data_impact_adherence)
model_diabetes_baseline_statin <- lm(medicine_adherence_statin ~ diabetes + baseline.condition, data = data_impact_adherence)

# Display model summary
summary_diabetes_baseline_ace <- summary(model_diabetes_baseline_ace)

summary_diabetes_baseline_bb <- summary(model_diabetes_baseline_bb)

summary_diabetes_baseline_statin <- summary(model_diabetes_baseline_statin)

# Function to create boxplots
create_boxplot <- function(medication) {
  ggplot(data_impact_adherence, aes(x = diabetes, y = get(paste0("medicine_adherence_", medication)))) +
    geom_boxplot() +
    labs(x = "Diabetes", y = "Adherence Rate", title = paste("Diabetes vs.", medication, "Adherence")) +
    theme_minimal() +
    facet_wrap(~baseline.condition, scales = "free_y")
}
```

##### Create boxplots for each medication

We can observe from the below box plots that:
1. Patients with a baseline condition of major heart attack/operation tend to be more adherent as compared to patients with a baseline condition of moderate symptoms or light procedure.
2. Patients with diabetes tend to be more adherent as compared to patients without diabetes

```{r box_plot_ace, warning=FALSE}
create_boxplot(ace.name)
```

```{r box_plot_bb, warning=FALSE}
create_boxplot(bb.name)
```

```{r box_plot_statin, warning=FALSE}
create_boxplot(statin.name)
```